Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2023-11-29 manu-draft.qmd
more discussion

Global increases in temperature and precipitation related to climate change have contributed to widespread alteration of the structure and functioning of natural environments.  At a local scale, these changes have been demonstrated herein for Tampa Bay where the evaluation of long-term trends in water temperature and salinity showed clear changes mirrored by long-term changes in air temperature and precipitation. Over the previous fifty years, the Bay has demonstrated a consistent trend towards hotter and fresher conditions, where water temperature has approximately increased by 0.03 - 0.07 $^\circ$C per year and salinity has decreased by 0.04 - 0.2 ppt per year.  These changes were demonstrated in three long-term datasets with different sampling methods and periods of record.  Understandably, the trends were most consistently observed in the EPC dataset with the longest period of record covering nearly fifty years of monthly observations, with the most dramatic changes occurring during the summer rainy season locations farther from the mouth of the Bay. These long-term changes manifested into consistent trends in the stressor metrics, where the continuous number of days each year when temperature was above, salinity was below, or both occurred showed consistent increases across the period of record. Notably, bottom and surface trends for water temperature and salinity in the EPC dataset were similar, suggesting that any potential stress on biotic resources related to these factors are likely occurring throughout the water column.

Similar regional, long-term changes in coastal waters have been observed by others. @nickerson2023 evaluated sea surface trends at a larger spatial scale for Tampa Bay, the West Florida Continental Shelf, and the adjacent Gulf of Mexico.  Temperature trends were similar for Tampa Bay (the EPC dataset was also used), although the authors note that the power of detecting the trends decreased for shorter time periods, as independently verified with the FIM and PDEM datasets herein.  @nickerson2023 also noted that temperature increases in Tampa Bay were most pronounced in the winter, which somewhat conflicts with our results showing distinct increases in the summer, particularly for the OTB and HB segments.  However, our assessment evaluated trends at individual EPC stations and bay segments, whereas @nickerson2023 evaluated the EPC data as an average for the entire bay for consistency of comparison to the larger spatial area.  These differences are understandable given the analysis goals and methods for data synthesis.  Our results showing notable increases in temperature and decreases in salinity for OTB and HB for the summer months  is likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico.  These uppper bay segments are more affected by hydrologic inflows (HB), lack of circulation (OTB), or thermal stress related to shallower depths. Regardless, @nickerson2023 state a Gulf-wide increase in water temperature of 0.1 - 0.5 $^\circ$C per decade, which is comparable to our estimates.

Our relatively simple modeling approach provides some evidence that climate-related stressors can explain the recent loss in seagrass in Tampa Bay. The models did not provide a consistent, nor statistically powerful, explanation that increasing temperature and decreasing salinity were key drives.  However, a weight-of-evidence approach evaluating all models together does suggest a pattern that demonstrates the value of considering multiple datasets and models in explaining noisy ecological patterns.

The seagrass loss in Tampa Bay since 2016 is a notable phenomenon that is not limited to our study area. Losses have been observed throughout southwest Florida during this time period, including... These regional losses suggest that large-scale stressors are driving these changes, both in Tampa Bay and elsewhere, leading us to our initial hypothesis that climate-related stressors could explain the change in Tampa Bay.     and

clearTakehome results: notable temp, salinity changes, some evidence of association with seagrass change. Of the limited evidence, effects of high temp, low salinity may just be additive, no support for interactions. This was shown by the FIM models and lack of significance for the "both" EPC model. Also, note that the PDEM data may have had low power, limited spatially (only west side of bay segment) and even shorter time period (supp fig, sal trends support this).

@@

Role of additional stressors and why they weren't considered in the model - they're important but we can also demonstrate that light attenuation has been relatively stable over time, so may not be important to include in the models.

Issues with thresholds, not definitive and those likely to stress seagrass are potentially not occurring at frequencies sufficient enough to model. Those used herein were chosen statistically. Point is that the temp and sal trends are in a direction such that thresholds that are more likely to stress seagrass are more likely to occur in the future.

Are we just showing seagrasses are not present because of temperature limitations or if temperature indeed caused a reduction?

@@

>>>>

Global increases in temperature and precipitation related to climate change have contributed to widespread alteration of the structure and functioning of natural environments.  At a local scale, these changes have been demonstrated herein for Tampa Bay where the evaluation of long-term trends in water temperature and salinity showed clear changes mirrored by long-term changes in air temperature and precipitation. Over the previous fifty years, the Bay has demonstrated a consistent trend towards hotter and fresher conditions, where water temperature has approximately increased by 0.03 - 0.07 $^\circ$C per year and salinity has decreased by 0.04 - 0.2 ppt per year.  These changes were demonstrated in three long-term datasets with different sampling methods and periods of record.  Understandably, the trends were most consistently observed in the EPC dataset with the longest period of record covering nearly fifty years of monthly observations, with the most dramatic changes occurring during the summer rainy season at locations farther from the mouth of the Bay. These long-term changes manifested into consistent trends in the stressor metrics, where the continuous number of days each year when temperature was above, salinity was below, or both occurred showed consistent increases across the period of record. Notably, bottom and surface trends for water temperature and salinity in the EPC dataset were similar, suggesting that any potential stress on biotic resources related to these factors are likely occurring throughout the water column.

Similar regional, long-term changes in coastal waters have been observed by others. @nickerson2023 evaluated sea surface trends at a larger spatial scale for Tampa Bay, the West Florida Continental Shelf, and the adjacent Gulf of Mexico.  Temperature trends were similar for Tampa Bay (the EPC dataset was also used), although the authors note that the power of detecting the trends decreased for shorter time periods, as independently verified with the FIM and PDEM datasets herein.  @nickerson2023 also noted that temperature increases in Tampa Bay were most pronounced in the winter, which somewhat conflicts with our results showing distinct increases in the summer, particularly for the OTB and HB segments.  However, our assessment evaluated trends at individual EPC stations and bay segments, whereas @nickerson2023 evaluated the EPC data as an average for the entire bay for consistency of comparison to the larger spatial area.  These differences are understandable given the analysis goals and methods for data synthesis.  Our results showing increases in temperature and decreases in salinity for OTB and HB for the summer months are likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico.  These upper bay segments are more affected by hydrologic inflows (HB), lack of circulation (OTB), or thermal stress related to more rapid warming with shallower depths. Regardless, @nickerson2023 state a Gulf-wide increase in water temperature of 0.1 - 0.5 $^\circ$C per decade, which is comparable to our estimates.

Our relatively simple modeling approach provided some evidence that climate-related stressors can explain the recent loss in seagrass in Tampa Bay. The models did not provide a consistent, nor statistically powerful, explanation that increasing temperature and decreasing salinity were key drivers.  However, a weight-of-evidence approach evaluating all models together does suggest a pattern that demonstrates the value of considering multiple datasets and models in explaining noisy ecological patterns. Model results for the EPC and FIM datasets both suggested that increasing temperature and decreasing salinity were associated with potential seagrass loss post-2016, described primarily using separate interaction terms of temperature or salinity with time period. For the EPC model, the interaction term was not significant for temperature, whereas the interaction was marginally significant for the salinity metric such that a negative association was observed post-2016 as compared to pre-2016 where frequency occurrence of seagrass declined with the increasing number of days when salinity was below 25 ppt. Likewise, the FIM models had a significant interaction term for the association of the temperature with time period and a marginally significant interaction term for the association of salinity with time period, such that percent cover of seagrasses showed a decline with increasing temperature and decreasing salinity post-2016.  An important distinction between the EPC and FIM models is that the former evaluated the number of days above/below thresholds each year to quantify increases in annual stress associated with temperature and salinity, whereas the latter evaluated observed temperature and salinity values at the time of seagrass sampling.  As such, both models attempted to describe the role of these stressors on potential seagrass change, but use different independent variables given the different sampling designs of each monitoring programs.  These differences highlight challenges describing noisy relationships in long-term ecological datasets, while also demonstrating the utility of our weight-of-evidence approach to describe these relationships.

An additional caveat of our models was the use of "thresholds" to define potential stressor metrics for temperature and salinity on an annual time scale. Our choice to use 30$^\circ$C and 25 ppt for temperature and salinity was primarily a statistical consideration given a consistent increase over time in the number of days when these thresholds were crossed, as shown by the mixed-effects linear models for each bay segment.  That is, sufficient change and variation in the independendent variables for the models of seagrass change were needed to statistically describe potential relationships. These thresholds are not to be considered ecologically significant since most of the reported values in tropical and sub-tropical environments suggest that the limits of the ecological niche for seagrasses are higher for temperature and lower for salinity. Because we did not see a dramatic increase in the number of days each year when the thresholds were crossed at more stressful values (i.e., higher temperature, lower salinity), conditions in Tampa Bay in recent years are generally within the ecological niche for seagrasses.  This does not suggest that these factors are unimportant, both currently and in the future.  Our models suggested that temperature and salinity are potentially associated with seagrass loss and the consistent long-term trend in temperature and salinity suggest that conditions are very likely to become more stressful in the futre.  Moreover, the GAMs used to model continuous temperature and salinity at each EPC monitoring station likely did not fully describe the current extent of stress given that extreme low or high values were often not captured by the daily predictions (see supplement).  Regardless of the above limitations, the role of climate stressors on seagrass change is likely to continue in the future as suggested by our results.

Additional limitations of our model may relate to an incomplete description of factors influencing seagrass growth.  This relates to the inclusion of additional drivers as well as issues related to correlation/causation.  For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of nutrients, nitrogen in particular, in affecting light environments for seagrass growth.  As such, light attenuation or water clarity could have been included in our models to more completely describe factors influencing growth, i.e., the residual differences could additionally be explained by temperature or salinity.  Our use of time period (pre-/post-2016) as a categorical variable partially addressed this issue.  Modeling seagrass change as related to temperature or salinity for the whole record without time period as a predictor would have shown a spurious correlation of both with seagrass given the long-term recovery of Tampa Bay related to nutrient reduction, increased water clarity, and increased seagrass coverage. Thus, time period was necessary to control for these confounding relationships.  Additionally, light attenuation  has been relatively consistent since 2016 and within the limits estimated to be supportive of seagrass growth in Tampa Bay (OTB in particular where the most losses have been seen). This suggests that other factors are influencing seagrass growth and the inclusion of light attenuation in the models was likely not critical for describing the roles of temperature and salinity. A final consideration for our models relates to how seagrasses may influence their environment, particularly for the EPC and PDEM datasets where temperature and salinity were measured at the same locations as seagrass. For example, temperature may simply be higher in locations where seagrasses are present, i.e., seagrasses may be influencing their environment rather than the environment influencing seagrasses.  This explanation cannot be ruled out with the existing datasets, although the trend analyses showing the consistent long-term changes and our weight-of-evidence modeling approach suggest that climate-related stressors driving seagrass loss is a more likely scenario.

The seagrass loss in Tampa Bay since 2016 is a notable phenomenon that is not limited to our study area. Losses have been observed throughout southwest Florida during this time period, including Sarasota Bay directly south of Tampa Bay and Charlotte Harbor further south. These regional losses suggest that large-scale stressors are driving these changes, leading us to our initial hypothesis that climate-related stressors could explain the change in Tampa Bay.  Based on our results, the losses elsewhere may potentially be explained by temperature and salinity and are worth exploring in other southwest Florida coastal regions and estuaries where long-term datasets exist. Additional factors that could explain these changes are also likely co-occurring with climate stress, some of which are unique to Tampa Bay and others that are likely more pervasive.  For Tampa Bay, annual summer/fall blooms of the toxic dinoflagellate *Pyrodinium bahamense* have occurred in OTB since 2008 and the specific relationships of these blooms with seagrass change is unclear, athough the expectation is that seagrass growth may be limited by the degradation of the light environment with algal growth.  These blooms are exacerbated by the hydrologic conditions in OTB that contribute to relatively longer residence times, whereas the effect of warming temperature and decreasing salinity in OTB will further complicate the understanding of how these blooms manifest and persist each year. In Tampa Bay and elsewhere, the persistence of macroalgal species and how these may inluence seagrass growth and recovery h as also been a concern...

@@

@@

### Conclusions
